%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% 		Dissertation template
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[a4paper, 12pt]{report}

\usepackage{import}
\usepackage{Utilities/utilities}

 
\input{Utilities/glossary.tex}


% define line-spacing
% \linespread{1.25}

% \renewcommand{\familydefault}{\sfdefault}
% \geometry{left=3cm,right=3cm,top=2cm,bottom=2cm}

% % % for proof read
% \linespread{1.75}

\begin{document}

\import{Utilities/}{title.tex}


%%%%%%%%%%% Table of Contents, Figures and Images %%%%%%%%%%% 

\setcounter{tocdepth}{4}
% \setcounter{tocdepth}{5}


%this needs to be defined before inserting the table of contents
% \setcounter{tocdepth}{1} % Show sections
% \setcounter{tocdepth}{2} % + subsections
% \setcounter{tocdepth}{3} % + subsubsections
% \setcounter{tocdepth}{4} % + paragraphs
% \setcounter{tocdepth}{5} % + subparagraphs


\pagenumbering{Roman}

\import{Sections/}{abstract.tex}
% \import{Sections/}{acknowledgments.tex}


\tableofcontents
\newpage

\listoffigures
\newpage

\listoftables
% \listoflistings

% \printglossary[type=acronym,style=long,nonumberlist]
\clearpage


\printglossary[nonumberlist]

\printglossary[type=\acronymtype,nonumberlist]
\newpage

\pagenumbering{arabic}
% \newpage



\import{Sections/}{intro.tex}

% \import{Sections/Lit_review/}{master.tex}
% \import{Sections/ClusteringAnalysis/}{master.tex}
% \import{Sections/Network_I/}{master.tex}
% \import{Sections/Network_I/EdgePruning}{master.tex}
\import{Sections/Network_II/}{master.tex}

\import{Sections/}{discussion.tex}


\chapter{Ethics statement}

There was no new data generated in this study and all the gene expression, mutational data was taken from existing resources.

Data for the Urotheliome project was generated by RNAseq from human urothelium either freshly isolated for in situ analysis (P0) or following cell culture. The tissues were collected at surgery with the following National Health Service (NHS) Research Ethics Committee (REC) approvals

\begin{itemize}
    \item REC reference: 99/095. "Use of excess ureter from transplant donor kidneys for research purposes". Leeds (East) Research Ethics Committee (latest annual report approved May 2023). These samples were collected as surgical waste and processed anonymously.
    \item REC reference: 21/YH/0225. "URoBank: Urothelial Research Tissue Bank (Phase 2)". Leeds (East) Research Ethics Committee (latest annual report approved Nov 2023). These samples were collected with consent and used anonymously either linked or unlinked to clinical records.  
\end{itemize}

Upon receipt, samples were ascribed a unique "Y" number (eg Y1423) during processing to separate the urothelial cells for RNAseq ("P0") and/or NHU cell culture.

\import{Sections/}{appendix.tex}


%%%%%%%% not include in the PhD %%%%%%%%
% \import{Sections/Gene_Sel/}{master.tex}
% \import{Sections/Miscellanious/}{master.tex}



\bibliographystyle{IEEEtranN}
\bibliography{paperpile}


\end{document}

