\subsection{A brief introduction}

This section will start with a brief overview of the Machine Learning field, with the emphasis on the work done in the last decade and what contributed to the hype. Will introduce the hard problems it solved and why \acrshort{dl} (\& its variations) represented a big step forward in the Machine Learning field. This will be followed by the limitations of this current approach that will introduce both \acrfull{snn} and \acrfull{ea}. Will go through some of their features, highlighting the important aspects such as \acrfull{es} or Neuron Models. After laying the foundation, concepts will need to present the work done by both EA \& SNN in biomedical research. 

Towards the end of this section, I want to do a comparison between the two computational approaches and maybe? the traditional Machine Learning. The goal will be to have a concise summary that will help us to make decisions in the future, rather than going through the entire literature. 

\begin{itemize}
\item Overview of Machine Learning school of thoughts (w/ history).  <-- One or two sentences for each to give a sense of the broadness in ML
  \begin{itemize}
    \item Bayesian
    \item Symbolists 
    \item Connectionists 
    \item Evolutionary 
  \end{itemize}
\item Methods that are milestones in the development of ML
    \begin{itemize}
      \item Multi-layer network 
      \item Back-propagation 
      \item Deep Learning 
      \item Convolution Neural Networks
      \item AlphaGo, Protein Folding etc. 
    \end{itemize}
\item The hard problems that ML solved
    \begin{itemize}
      \item CV
      \item Speech to text
      \item Translation
      \item Go 
    \end{itemize}
\item Explainable AI - I thinking it’s worth dedicating a paragraph to it
\item Showing that there are some limitations to ML
\begin{itemize}
  \item The fail promises of AI mainly refering to self-driving
  \item find it interesting to talk about the overly optimistic predictions of AI and how the DL trend contributed to that
  \end{itemize}
\end{itemize}

