\subsection{Networks in molecular biology} \label{s:lit:nets_bio}

\vspace{3mm}
% \noindent\rule{17cm}{0.2pt}
\fbox {
    \parbox{\linewidth}{
      \begin{itemize}
        \item Co-expressed networks
        \item Integrative Network approaches
        \item Other network approaches
      \end{itemize}
    }
}
\vspace{3mm}

% Make the case for pathways and gene interactions - Why it is important to study them
Interactions and the relationships between multiple components are central to tissue functioning and biological processes. There are multiple levels of interactions in molecular biology which are represented by gene pathways where one gene influences another to produce proteins, and that in turn produces more proteins until that pathway fulfils its function. When certain components of cellular mechanisms fail, such as a mutation in a gene that plays a critical role in key pathways (e.g., those involved in the cell cycle), the immediate impact on the tissue may not be apparent. Over time, however, the effects of the mutation can accumulate, influencing other mechanisms and potentially leading to a cascade of biological disruptions. This inter-connectivity between genes and cellular processes underpins the widespread use of networks in genomics, which has facilitated rapid advancements in the field.

% Need to address the network pathways 
The relevant network concepts were introduced in \cref{s:lit:graph_overview}
and how graphs were used to analysed gene pathways is covered in \cref{s:lit:multi-view}. However, the information about pathways is still being researched and although a complete representation of a tissue is not available, there are other alternatives to analyse the gene interactions in a tissue.

% Why are we choosing the co-expressed networks?
One approach is the co-expressed networks which uses different correlation metrics (e.g. Spearman, Pearson or partial-correlation) from gene expression to model the strengths between the nodes (edges). The rationale behind it is that genes expressed together have a higher correlation and subsequently have a stronger connection between nodes. Conversely, genes with a weaker connection are represented are lower correlated. After the network is built, related genes are grouped together in communities which are then used to subtype the disease. These network approaches are covered in Section \ref{s:lit:co_net} and are crucial for understanding the methods developed in this PhD project. The co-expressed networks are the preferred network approach in this project as these are simple, versatile and meet the aims and objectives of this project outlined in \cref{s:lit:aims_objs}. It also offers a more convenient and adaptable method to stratify the diseases on multiple sources of omics.

% DEA as an alternative
Another alternative to co-expressed networks is to build the network based on the Deferentially Expressed Analysis (DEA) data between the healthy tissue and tumour. While slightly more restrictive, this approach will highlight the differences between the two tissue states and potentially provide new biological insights. The review \citet{Van_Dam2018-id} covers such network approaches and also acts as a guide on building DEA-based networks. Using DEA between healthy tissue and tumours has been considered in the project but the co-expressed network was preferred for its versatility.

% Alternative methods for co-expressed networks - in the review from Petti
The review from 2023 of \citet{Petti2023-qo} covers the different network methods used in the field to stratify a disease, some of which are explored in this project. The patient similarity network section covers different methods using co-expression networks but not the work of \citet{Care2019-ij} on PGCNA (2019) which is one of the main sources in this project. Nevertheless, the review offers good insights on alternatives such as the network-based stratification (NBS), initially proposed for integrating data by \citet{Hofree2013-ld}, covered in \cref{s:lit:net_prop}, or the multi-layer (\cref{s:lit:multi-layer}) networks.

This section covers the work on co-expressed networks with a focus on PGCNA and WGCNA as well as research done using partial-correlation as an alternative to the standard methods (Spearman, Pearson) to build a graph. The networks constructed using a Bayesian approach is a topic dealth with in the next section. \Cref{s:lit:net_data_int} provides an overview of network propagation and multi-layer graphs as methods to integrate multiple data types. Lastly, the \cref{s:lit:comm_detect} is dedicated to community detection algorithms as they have a fundamental role in finding the groups of genes with a biological function.

\import{Sections/Lit_review/}{co-expressed_net.tex}

\import{Sections/Lit_review/}{comm_detection.tex}





