\thispagestyle{plain}
\begin{center}
    \Large        
    \textbf{Acknowledgements}
    \vspace{0.9cm}

\end{center}
 

My four years as a PhD student were not just a period of academic growth, but also a time of personal transformation. This growth was made possible by the unwavering support and guidance of my family, Diana, friends, and supervisors. Your support has helped me not only academically but also personally.

Doing a PhD is never an individual effort, even more so in an interdisciplinary project. I was fortunate to have a fantastic team of supervisors who helped me navigate the computational and biology realms and the PhD journey itself. Thank you for your support, feedback and lessons throughout the years. Steve, thank you for being instrumental in setting up the project, Jenny for welcoming me to the JBU team and supporting my learning of urothelium, and David for all the invaluable computational insights. Andrew, your guidance in research and mentorship in the academic world were vital to my project and how I have grown; thank you. I hope to one day pay this forward and inspire others in the same way.

The generous funding from the Departments of Electronics and Biology and York Against Cancer (YAC) facilitated the success of my project. I am deeply grateful for the teaching opportunities the Department of Electronics provided, which significantly contributed to my academic formation.

The four years of my PhD represented an exciting academic time during which I met some extraordinary people who shaped my experience. My electronics PhD fellows Vignesh, Cameron, and James, and my colleagues from the Jack Birch Unit, I learnt a great deal from you, and thank you for being around my journey. I also want to extend my appreciation for my friends outside of academia, Vlad and my cousin Matei, who patiently listened to my PhD experience when it was not at its highest.

While the PhD represented 'only' four years of my academic life, it would not have been possible without my family's unwavering support and love. Dad, Mom, Ion, and Fany, your decision to send me for my undergraduate studies in the UK profoundly impacted who I became as an individual. Mihnea, your constant support as my brother and best friend is priceless. Thank you all for being there and supporting me throughout my life.

Diana, we have both been part of each other's PhD journeys, and I could not have chosen a better partner to have during these times. We both survived! Thank you for everything, for bringing light and clarity into my life and ensuring I am not overworking myself. Thank you for all the memories in these four years, and I can't wait to embark on our next journey together!











\newpage 

\thispagestyle{plain}
\begin{center}
    \Large        
    \textbf{Author's declaration}
    \vspace{0.9cm}

\end{center}


I declare that this thesis is a presentation of original work and I am the sole author. This work has not previously been presented for a degree or other qualification at this University or elsewhere. All sources are acknowledged as references.
